# From theory to model: (generalized) linear models {#sec-theory-to-glm}

The previous chapter drew the chain from research question to interpretation, but left one link abstract: where does the *model* actually come from? This chapter builds that link, step by step, for the most common case in the social and health sciences: a verbal theory, observational data, and (generalized) linear models. The approach follows a tradition best set out by @saris1984: treat the theory as the object of interest, and the statistical model as its translation — precise enough to estimate, precise enough to be wrong.

By the end of the chapter you will have gone from a paragraph of prose to estimated equations and back: the estimates saying something about the world (*if* the model is true), and the tests saying something about the theory.

## A verbal theory

Here is a small theory of the kind you might find in the occupational health literature:

> Demanding jobs cause stress. Having control over one's own work protects against stress. Stress, in turn, causes health complaints; job demands affect health complaints only through stress.

This is respectable science prose, and it is *not yet* a model. Try to estimate "demanding jobs cause stress" and you will notice everything that is missing: how strong is the effect? Is it the same for everyone? What else causes stress? What should we measure?

**Step 1: distill the claims.** Rewrite the theory as an explicit list of directed claims, one per pair of concepts:

1. Job demands ($D$) increase stress ($S$): positive effect, $D \rightarrow S$.
2. Autonomy ($A$) decreases stress: negative effect, $A \rightarrow S$.
3. Stress increases health complaints ($C$): positive effect, $S \rightarrow C$.
4. Demands have **no direct effect** on complaints: no arrow $D \rightarrow C$.

Claim 4 deserves a pause. The words "only through stress" — easy to skim in prose — became a claim of *absence*. Absences are the most scientifically valuable claims a theory can make, because they are the easiest to prove wrong. A theory that merely says "everything affects everything, sign unknown" can never be embarrassed by data.

We must also decide what to say about relationships the theory is silent on: may demands and autonomy be related to each other? We do not wish to explain that relationship, only to allow it (demanding jobs plausibly come with less autonomy). So we let $D$ and $A$ covary freely, without a directed claim.

::: {.callout-tip icon="false" title="Exercise 2.1"}
Distill this verbal theory into directed claims, including any claims of absence: "Loneliness undermines sleep. Poor sleep worsens depressive symptoms, and so does loneliness itself, even apart from its effect on sleep. Physical activity improves sleep but has no further connection to depressive symptoms."
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.1"}
With loneliness $L$, physical activity $P$, sleep quality $Q$, depressive symptoms $Y$: (1) $L \rightarrow Q$, negative. (2) $Q \rightarrow Y$, negative. (3) $L \rightarrow Y$, positive, direct (the "even apart from" clause). (4) $P \rightarrow Q$, positive. (5) Absence: no direct $P \rightarrow Y$ (only via sleep). $L$ and $P$ are both exogenous; the theory says nothing about their relationship, so we allow them to covary freely. Note how "no further connection" became a testable claim of absence.
:::

## From claims to equations

**Step 2: write each endogenous variable as an equation.** A variable that receives at least one arrow is called *endogenous* (explained within the model); the others are *exogenous* (taken as given). Each endogenous variable gets one linear equation:

$$
\begin{aligned}
S &= \beta_1 D + \beta_2 A + \varepsilon_S \\
C &= \beta_3 S + \varepsilon_C
\end{aligned}
$$ {#eq-stress-model}

Three features of these equations carry all the content:

*The coefficients.* $\beta_1$ is the number of units $S$ changes when $D$ increases by one unit *and $A$ is held constant*. The theory's claims become sign constraints: $\beta_1 > 0$, $\beta_2 < 0$, $\beta_3 > 0$. The claim of absence is visible as a missing term: there is no $D$ in the equation for $C$.

*The error terms.* $\varepsilon_S$ collects everything else that affects stress: genes, family, the neighbor's renovation. Including it is not an admission of defeat but part of the theory's honesty: we claim demands and autonomy matter, not that they are everything. Crucially, we assume each error is **uncorrelated with the predictors in its equation**. That is the observational-study assumption from @sec-rq-to-interpretation wearing mathematical clothes: it says the omitted causes of stress do not travel together with demands and autonomy. No amount of data can verify it; the design (or your defense of it) must carry it.

*What is absent.* No squared terms, no products: effects are assumed **linear** (each extra unit of $D$ does the same to $S$ as the previous one) and **additive** (the effect of $D$ does not depend on the level of $A$). The verbal theory said nothing either way; translating it forced a choice. We choose the simplest form as a working assumption — and we should remember we made it.

To keep the arithmetic transparent, assume in this chapter that all variables are **standardized**: mean 0, standard deviation 1. Coefficients are then correlations-scale quantities, and a picture becomes possible.

**Step 3: draw the model.** The same model as a *path diagram*:

```{mermaid}
%%| fig-cap: "Path diagram of the stress model. Single-headed arrows are causal claims; the double-headed arc is a covariance we allow but do not explain; the epsilons are everything-else."
flowchart LR
  D[Demands] -->|β₁ +| S[Stress]
  A[Autonomy] -->|β₂ −| S
  S -->|β₃ +| C[Complaints]
  D <-.->|r| A
  eS((ε_S)) --> S
  eC((ε_C)) --> C
```

A path diagram is not an illustration *of* the model; it *is* the model, in graphical notation. Every arrow is a coefficient, every missing arrow a claim of absence, every double-headed arc an unexplained covariance.

## What the model implies: the tracing rules

Here is where the translation pays off. Once the theory is a set of linear equations, it makes *quantitative predictions* about things we can observe — in particular, about correlations. For standardized models without feedback loops, the predictions follow from **Wright's tracing rules** [@wright1934]:

> The model-implied correlation between two variables is the **sum over all valid traces** connecting them, where each trace contributes the **product of its coefficients**. A valid trace (1) never passes through the same variable twice, (2) never goes forward along arrows and then backward, and (3) uses at most one double-headed arc.

You may move *backward* along arrows first (against the arrowheads), optionally cross one double-headed arc, and then move *forward*; once moving forward, there is no turning back.

Let us trace the implied correlation between demands and stress, writing $r$ for the $D$–$A$ correlation:

- Trace 1: $D \rightarrow S$, contributing $\beta_1$.
- Trace 2: $D \leftrightarrow A \rightarrow S$, contributing $r\,\beta_2$.

$$
\rho_{DS} = \beta_1 + r\beta_2.
$$

And between demands and complaints:

- Trace 1: $D \rightarrow S \rightarrow C$: $\beta_1 \beta_3$.
- Trace 2: $D \leftrightarrow A \rightarrow S \rightarrow C$: $r\,\beta_2\beta_3$.

$$
\rho_{DC} = \beta_1\beta_3 + r\beta_2\beta_3 = \rho_{DS}\,\beta_3.
$$

Notice what just happened: the theory now predicts a *relationship between correlations*. Whatever the values of the coefficients, this model requires $\rho_{DC} = \rho_{DS} \times \rho_{SC}$ (since $\rho_{SC} = \beta_3$; check it by tracing). If the data show, say, a strong demands–complaints correlation that survives when stress is held constant, the model — and with it claim 4 of the theory — is in trouble. The claim of absence became a testable prediction. This single idea, that a causal structure implies (and is tested through) constraints on observable correlations, is the foundation on which all of structural equation modeling rests, and we will lean on it hard in @sec-identification.

::: {.callout-tip icon="false" title="Exercise 2.2"}
Using the tracing rules, derive the implied correlations $\rho_{AS}$, $\rho_{SC}$, and $\rho_{AC}$ for the stress model. Then compute all five implied correlations numerically for $\beta_1 = 0.4$, $\beta_2 = -0.3$, $\beta_3 = 0.5$, $r = -0.3$.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.2"}
$\rho_{AS} = \beta_2 + r\beta_1$ (traces $A \rightarrow S$ and $A \leftrightarrow D \rightarrow S$). $\rho_{SC} = \beta_3$ — the only valid trace is $S \rightarrow C$; a trace like $S \leftarrow D \leftrightarrow A \rightarrow S \rightarrow C$ visits $S$ twice and is invalid. $\rho_{AC} = \rho_{AS}\beta_3 = (\beta_2 + r\beta_1)\beta_3$. Numerically: $\rho_{DS} = 0.4 + (-0.3)(-0.3) = 0.49$; $\rho_{AS} = -0.3 + (-0.3)(0.4) = -0.42$; $\rho_{SC} = 0.5$; $\rho_{DC} = 0.49 \times 0.5 = 0.245$; $\rho_{AC} = -0.42 \times 0.5 = -0.21$. Note $\rho_{DC} = \rho_{DS}\rho_{SC}$, the model's testable constraint.
:::

## Estimating the model

Time to meet data. We simulate a dataset that behaves exactly according to the model — so we know the truth — and estimate the coefficients by regression, one equation at a time.

```{r}
#| label: simulate-stress
set.seed(2)
n <- 800
r <- -0.3
# Exogenous variables with correlation r:
D <- rnorm(n)
A <- r * D + sqrt(1 - r^2) * rnorm(n)
# Structural equations (standardized coefficients):
b1 <- 0.4; b2 <- -0.3; b3 <- 0.5
var_eS <- 1 - (b1^2 + b2^2 + 2 * b1 * b2 * r)   # so that Var(S) = 1
S <- b1 * D + b2 * A + rnorm(n, sd = sqrt(var_eS))
C <- b3 * S + rnorm(n, sd = sqrt(1 - b3^2))

stress <- data.frame(D, A, S, C)
round(coef(lm(S ~ D + A, data = stress)), 2)
round(coef(lm(C ~ S, data = stress)), 2)
```

The estimates land close to the true values $(0.4, -0.3, 0.5)$, as they should. In real data we would not know the truth; the estimates would *be* our knowledge, and their meaning is fixed by the model:

> **Interpretation of a linear regression coefficient.** A one-unit increase in $D$ is associated with a $\hat\beta_1$-unit increase in $S$, holding the other predictors in the equation constant. Under the model's assumptions — linearity, additivity, and errors uncorrelated with predictors — "associated with" strengthens to "leads to."

The same numbers, two readings. The weak reading (association) needs almost no assumptions and answers almost no interesting question. The strong reading (effect) answers the theory's question and needs every assumption in the model. There is no analysis-only trick that upgrades one to the other; that upgrade is purchased — or not — by the design and the defense of the error assumption.

We can also estimate all equations at once, which will matter as soon as we care about the system as a whole:

```{r}
#| label: lavaan-stress
library(lavaan)
model <- "
  S ~ D + A
  C ~ S
"
fit <- sem(model, data = stress)
coef(fit)[1:3]
```

And now the payoff of claims of absence: the model *forbids* a direct $D \rightarrow C$ effect, so it claims the restriction can be imposed without harming the fit to the data. That is testable:

```{r}
#| label: test-absence
fit_free <- sem("S ~ D + A \n C ~ S + D", data = stress)
anova(fit, fit_free)
```

The test compares our restricted model against one where demands may also affect complaints directly. A small, nonsignificant chi-square difference says the data are content with the absence — as they should be here, since we built the data that way. In real research this is the moment the theory risks something: had the test rejected, claim 4 would have been falsified (assuming the rest of the model!). A test never confronts one claim with the data alone; it confronts one claim with the data *given all the other claims*. And even when the whole model fits, it is not thereby proven true — other models fit equally well, a problem so important it gets its own chapter (@sec-identification).

::: {.callout-tip icon="false" title="Exercise 2.3"}
Predicted values follow from the fitted equations. Using the true coefficients ($\beta_1 = 0.4$, $\beta_2 = -0.3$, $\beta_3 = 0.5$): (a) What stress level does the model predict for a person one standard deviation above average in demands and one below average in autonomy? (b) What complaint level does it predict for that same person? (c) Why is your answer to (b) *not* obtained by plugging the person's actual stress — which you do not know — into the second equation?
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.3"}
(a) $\hat S = 0.4(1) - 0.3(-1) = 0.7$: 0.7 standard deviations above average stress. (b) Chain the equations: $\hat C = 0.5 \times \hat S = 0.35$. (c) Because prediction from $D$ and $A$ alone must route through the *predicted* stress; the person's actual stress also contains $\varepsilon_S$, which is unknown. If we *did* know actual stress, the model says demands and autonomy tell us nothing more ($C$'s equation contains only $S$) and we would predict $0.5 \times S$ directly.
:::

## When the outcome is yes/no: logistic regression

Suppose the outcome is not a complaints scale but an event: whether the employee goes on long-term sick leave ($Y = 1$) or not ($Y = 0$). A linear equation is now awkward — it happily predicts probabilities of $-0.2$ or $1.4$. The standard repair keeps the linear machinery but changes what the equation predicts: not the probability itself, but its **log-odds**.

Odds are a currency for probability used by bookmakers: probability $p$ corresponds to odds $p/(1-p)$. A probability of $0.75$ is odds of $3$ ("three to one"); $0.5$ is odds of $1$. Taking the logarithm stretches this scale over the whole number line: probability $0 \to$ log-odds $-\infty$, probability $1 \to +\infty$, probability $0.5 \to 0$. On the log-odds scale, a linear equation can roam freely and every value maps back to a legal probability:

$$
\log \frac{P(Y=1)}{1 - P(Y=1)} = \alpha + \beta S .
$$

This is **logistic regression**, the second member of the family of *generalized* linear models we will use (the first was linear regression itself: same idea, identity instead of log-odds). The price of legality is paid in interpretation:

> **Interpretation of a logistic regression coefficient.** A one-unit increase in $S$ leads to a $\beta$-unit increase in the *log-odds* of $Y$ — under the same kinds of assumptions as before.

Nobody thinks natively in log-odds, so two conversions are essential. First, exponentiating gives the **odds ratio**: each unit of $S$ *multiplies the odds* by $e^{\beta}$. A useful rule of thumb: for smallish coefficients (roughly $|\beta| < 0.3$), $e^{\beta} \approx 1 + \beta$, so the odds increase by about $100 \times \beta$ percent. A coefficient of $0.1$ raises the odds by about 10.5%; the shortcut says 10%.

Second, for any concrete person we can compute an actual **probability** by inverting the log-odds:

$$
P(Y = 1) = \frac{1}{1 + e^{-(\alpha + \beta S)}} .
$$

```{r}
#| label: logistic-stress
set.seed(3)
alpha <- -2; beta <- 0.8
p <- plogis(alpha + beta * S)         # inverse log-odds
Y <- rbinom(n, 1, p)
fit_log <- glm(Y ~ S, family = binomial)
round(coef(fit_log), 2)
round(exp(coef(fit_log)["S"]), 2)      # odds ratio per SD of stress
round(plogis(sum(coef(fit_log) * c(1, 1))), 2)  # P(sick leave) at S = 1
```

Read the output the way the chapter has taught you to read everything: a person one standard deviation higher in stress has $e^{0.69} \approx 2.0$ times the estimated odds of long-term sick leave (the truth we built in was $e^{0.8} \approx 2.2$); a person at $S = 1$ has a predicted probability around $0.20$, against about $0.11$ at average stress. And all of it means what it appears to mean only under the model's assumptions — which the design, not the arithmetic, must justify.

One warning specific to the log-odds scale: because the effect on the *probability* depends on where you start (the same odds ratio moves $p = 0.5$ much more than $p = 0.01$), always report what a coefficient means for realistic values of the predictors, not only the odds ratio.

::: {.callout-tip icon="false" title="Exercise 2.4"}
A logistic regression of hospital readmission on frailty score (standardized) gives $\hat\alpha = -1.5$, $\hat\beta = 0.4$. (a) Interpret $\hat\beta$ on the log-odds scale, exactly. (b) Give the odds ratio and check the rule of thumb. (c) Compute the predicted readmission probability for patients at average frailty and at two standard deviations above average. (d) A colleague summarizes: "each SD of frailty raises readmission probability by 40%." Correct them precisely.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.4"}
(a) Each additional standard deviation of frailty increases the log-odds of readmission by 0.4, other things (there are none here) equal. (b) $e^{0.4} = 1.49$: odds multiply by about 1.5. The rule of thumb says $1 + 0.4 = 1.4$ — usable but rough at $\beta = 0.4$, which is at the edge of the rule's range. (c) At average frailty: $\text{plogis}(-1.5) = 0.18$. At $+2$ SD: $\text{plogis}(-1.5 + 0.8) = \text{plogis}(-0.7) = 0.33$. (d) The 0.4 lives on the log-odds scale, not the probability scale: odds — not probability — increase by roughly 49% per SD. In probability terms the effect depends on the starting point; here, going from average to $+1$ SD moves the probability from 0.18 to about 0.25.
:::

## Why only linear models here?

Readers who have heard of random forests or neural networks may feel a growing unease: real relationships are not all linear-and-additive, so why is this chapter pretending they are? Three honest reasons.

First, nothing about the chain *question → design → assumptions → model → interpretation* is specific to linear models. Flexible, nonparametric methods relax the linearity and additivity assumptions, but they answer research questions through a design and a model in precisely the same way — the logic is just harder to see under the extra mathematics and computation. Learn the logic where it is transparent.

Second, in the modest sample sizes typical of social and health research, the data often cannot support much more than linear-with-a-few-terms anyway: flexibility must be paid for with data, and the linear model is frequently all we can afford — a decent local approximation estimated precisely, rather than the true curve estimated as noise.

Third, when you do have the data for it — large samples, many interactions, strong nonlinearities, prediction as the goal — flexible methods are the right tool, and later chapters use them. When you get there, bring this chapter along: those methods, too, produce answers only through assumptions that some design must make plausible.

## Test yourself {.unnumbered}

::: {.callout-tip icon="false" title="Exercise 2.A"}
Consider this verbal theory: "Neighborhood deprivation harms adolescents' school performance in two ways: it reduces access to quiet study space, and it increases exposure to stressors. Stressor exposure harms performance only by disrupting sleep. Study space has no relationship with sleep."

(a) Distill the theory into directed claims, including all claims of absence.
(b) Define the variables, state which are exogenous, and write the structural equations (standardized), including error terms.
(c) Draw the path diagram (paper and pencil suffice).
(d) Using the tracing rules, derive the model-implied correlation between deprivation and performance, in terms of your coefficients.
(e) State one constraint on observable correlations that could falsify the "only by disrupting sleep" claim.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.A"}
(a) Directed claims: deprivation reduces study space (−); deprivation increases stressor exposure (+); stressor exposure disrupts sleep (−); sleep improves performance (+); study space improves performance (+). Claims of absence: no direct stressors → performance ("only by disrupting sleep"); no path between study space and sleep ("no relationship"); no direct deprivation → sleep or deprivation → performance (deprivation works only the two named ways). (b) Let deprivation $N$ (exogenous), study space $Q$, stressor exposure $X$, sleep $Z$, performance $Y$:
$Q = \gamma_1 N + \varepsilon_Q$ ($\gamma_1 < 0$); $X = \gamma_2 N + \varepsilon_X$ ($\gamma_2 > 0$); $Z = \gamma_3 X + \varepsilon_Z$ ($\gamma_3 < 0$); $Y = \gamma_4 Q + \gamma_5 Z + \varepsilon_Y$ ($\gamma_4 > 0, \gamma_5 > 0$). (c) Five boxes, five single-headed arrows, an error on each endogenous variable — check your diagram against the equations: every arrow a coefficient, every missing arrow a claim of absence. (d) Two valid traces connect $N$ and $Y$: $N \rightarrow Q \rightarrow Y$ and $N \rightarrow X \rightarrow Z \rightarrow Y$, so $\rho_{NY} = \gamma_1\gamma_4 + \gamma_2\gamma_3\gamma_5$. (e) The model implies that the partial association between $X$ and $Y$, holding $Z$ and $N$ constant, is zero: stressors carry no information about performance once sleep is held constant (and deprivation too, since deprivation connects $X$ to $Y$ through the study-space path). A clearly nonzero $X$–$Y$ association given $Z$ and $N$ falsifies the "only by disrupting sleep" claim — conditional, as always, on the rest of the model.
:::

::: {.callout-tip icon="false" title="Exercise 2.B"}
A researcher fits the stress model of this chapter to real data from 600 employees and obtains $\hat\beta_1 = 0.35$, $\hat\beta_2 = -0.05$ (n.s.), $\hat\beta_3 = 0.45$, and the chi-square test of the omitted $D \rightarrow C$ path rejects ($p < 0.001$), with the freed direct effect estimated at $0.30$.

(a) What do these results say about each of the theory's four claims?
(b) The researcher concludes: "autonomy does not protect against stress." Give two distinct reasons this conclusion overreaches, one statistical and one design-based.
(c) The researcher adds the direct path $D \rightarrow C$ and reports the revised model as confirmed by the same data. Which idea from @sec-rq-to-interpretation does this violate, and what would be the sound way to proceed?
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.B"}
(a) Claim 1 (demands raise stress) is supported in sign and size; claim 2 (autonomy protects) is not supported — the estimate is near zero; claim 3 (stress raises complaints) is supported; claim 4 (no direct effect) is falsified given the rest of the model: demands appear to affect complaints partly outside the stress path. (b) Statistical: a nonsignificant coefficient is not evidence of absence; the confidence interval may include protective effects of practical size (absence of evidence ≠ evidence of absence). Design-based: the error-term assumption may fail — e.g., people prone to stress may select into (or report) low-autonomy jobs, biasing $\hat\beta_2$ toward zero; or autonomy may be measured with error, attenuating its coefficient. (c) It converts an exploratory finding into a pretend-confirmatory one: the revised model was chosen *because* it fit these data, so the "confirmation" is circular. Sound: report the modification as exploratory and test the revised model on new data (or a held-out portion collected for that purpose).
:::

::: {.callout-tip icon="false" title="Exercise 2.C"}
A study of smoking cessation reports a logistic regression of quitting ($Y = 1$) on number of counseling sessions attended ($X$, range 0–10): $\hat\alpha = -2.2$, $\hat\beta = 0.25$.

(a) Interpret $\hat\beta$ exactly, and via the odds-ratio rule of thumb.
(b) Compute predicted quitting probabilities at 0, 4, and 8 sessions. Comment on how the *probability* change per session differs across that range, even though $\beta$ is constant.
(c) An enthusiast concludes each additional session causes a 28% increase in the odds of quitting, so mandating 10 sessions would maximize quitting. Using the chain from @sec-rq-to-interpretation, identify the two distinct inferential leaps and the assumption each requires.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 2.C"}
(a) Each additional session increases the log-odds of quitting by 0.25; odds ratio $e^{0.25} = 1.28$; rule of thumb: $\approx +25\%$, close to the exact 28%. (b) $\text{plogis}(-2.2) = 0.10$; $\text{plogis}(-2.2 + 1) = 0.23$; $\text{plogis}(-2.2 + 2) = 0.45$. From 0→4 sessions: +0.13; 4→8: +0.22 — the same odds ratio moves probability more near 0.5 than near 0. (c) Leap 1, association to causation: requires that attendance is as-good-as-random given the model — but motivated quitters attend more sessions (confounding); the design (observational attendance) does not support it. Leap 2, effect to policy: even granting causation among voluntary attenders, *mandating* sessions intervenes on a different variable (compulsory attendance) for a different population (including the unmotivated); extrapolating the coefficient to 10 sessions also assumes linearity on the log-odds scale to the end of the range. Each leap needs its own design — the second, ideally, an RCT of mandated counseling.
:::
