# Machine learning without a teacher: clustering and mixture models {#sec-unsupervised}

The machine learning of the last three chapters was **supervised**: every observation in the experience carried the outcome, the recorded outcomes supervised the fit, and the performance measure compared predictions against outcomes the model had not seen. This chapter removes the supervisor. In **unsupervised learning** there is no outcome variable: the data are a set of observations on some features, and the task is to find structure among them, most commonly *groups* of observations that resemble each other. Return to Mitchell's definition from @sec-overfitting and notice what breaks. The task and the experience survive: find groups, using these data. The performance measure does not, because there is no recorded outcome against which a proposed grouping can be scored. The judge that chapters 6 and 7 relied on has left the room, and the central question of this chapter is what to do about that.

Social and health scientists want hidden groups for several recurring reasons. Sometimes the variable of real interest was never measured, but other variables carry information about it: how students answer exam questions is indicative of whether they have mastered the material, and how a chat partner responds is indicative of whether they are human or a bot. Sometimes an effect is suspected to differ between kinds of people, without the kinds being directly observed. And sometimes there are simply too many variables to interpret, and the analyst wants to reduce them to a few interpretable groups, as marketing does when it builds "segments". The techniques of this chapter serve all three purposes, and they come in two families that the chapter treats in turn: **clustering algorithms**, which group observations by their distances to each other, and **mixture models**, which posit an explicit statistical model in which an unobserved, discrete variable produces the groups.

## Two kinds of eruption

The Old Faithful geyser in Yellowstone erupts every 35 to 120 minutes, and each eruption lasts between one and five minutes. The `faithful` dataset records 272 eruptions on two features: the duration of the eruption and the waiting time since the previous one. @fig-faithful plots them.

```{r}
#| label: fig-faithful
#| fig-cap: "Old Faithful: duration of 272 eruptions against the waiting time since the previous eruption. Two kinds of eruption suggest themselves: short ones after a short wait, and long ones after a long wait."
#| fig-width: 5
#| fig-height: 3.6
plot(faithful$eruptions, faithful$waiting,
     xlab = "eruption duration (min)", ylab = "waiting time (min)")
```

Nobody labelled these eruptions, yet the picture shows two groups: short eruptions after short waits, and long eruptions after long waits. The eye performs unsupervised learning effortlessly in two dimensions. The methods of this chapter exist because real problems have ten or a thousand dimensions, where the eye fails, and because even in two dimensions the eye cannot say *how many* groups there are when the picture is less obliging than this one.

## Clustering by distance

The algorithmic family starts from a simple idea: a cluster is a set of observations that are close together, so the raw material is the **distance** between every pair of observations. For continuous features the default is Euclidean distance, the ordinary straight-line distance in feature space, but the choice matters and is wider than it first appears. Before any distance is computed, the features should usually be put on a comparable scale, typically by **standardization**: subtracting each feature's mean and dividing by its standard deviation. Otherwise a feature measured in centimetres contributes a hundred times more to the distance than the same information measured in metres, and the clustering quietly becomes a clustering on whichever feature has the largest numbers. And beyond continuous features, distances have been defined for almost any kind of object: edit distance for text and DNA sequences, dynamic time warping and Fréchet distance for time series and curves, shortest-path distances for networks. Any method that accepts a distance matrix can therefore cluster any kind of object for which a distance can be defined, which is a large part of clustering's flexibility.

**Hierarchical clustering** uses the distance matrix directly. The agglomerative version starts with every observation as its own cluster, repeatedly merges the two most similar clusters, and stops when only one cluster remains; the analyst then chooses the number of clusters afterwards, by cutting the resulting tree at some height. How similar two *clusters* are, given the distances between their *members*, is a choice called the **linkage**: complete linkage uses the largest distance between members, single linkage the smallest, average linkage the mean. The linkage is a hyperparameter, and different choices can produce quite different trees from the same data.

```{r}
#| label: fig-dendro
#| fig-cap: "Dendrogram of hierarchical clustering (complete linkage) on the standardized Old Faithful data. The height at which two branches join is the distance at which their clusters merged; cutting the tree at the dashed line yields two clusters."
#| fig-width: 7
#| fig-height: 3.2
d_f  <- dist(scale(faithful))
hc_f <- hclust(d_f, method = "complete")
plot(hc_f, labels = FALSE, hang = -1, main = "", xlab = "", sub = "", ylab = "merge height")
abline(h = 4, lty = 2)
```

@fig-dendro shows the result for the standardized Old Faithful data as a **dendrogram**: a tree in which the height at which two branches join is the distance at which the corresponding clusters were merged. Long vertical stretches without merges indicate well-separated clusters, and cutting the tree at the dashed height produces the two clusters the scatterplot suggested. The dendrogram displays every number of clusters at once, from $n$ down to one, which is informative but also a reminder that the method itself never decides how many clusters there are. Because hierarchical clustering needs the full $n \times n$ distance matrix, it becomes expensive when $n$ is large.

**K-means** takes the opposite route: fix the number of clusters $K$ in advance, and search for the partition of the observations into $K$ clusters that minimizes the total squared Euclidean distance between each observation and the mean of its cluster, its **centroid**. The exact search is infeasible, since the number of possible partitions is astronomical, so the standard algorithm iterates two steps from a random starting assignment: assign each observation to its nearest centroid, then recompute each centroid as the mean of its assigned observations, and repeat until nothing changes. Each step lowers the criterion, so the algorithm always converges, but only to a *local* minimum that depends on the random start. Two consequences follow for practice. The labels of the clusters are arbitrary, so cluster 1 in one run can be cluster 2 in the next, a phenomenon called **label switching** that also matters for the mixture models later in this chapter. And a single run can simply find a poor solution, so the algorithm should always be run from many random starts, keeping the best result; in R, `kmeans(x, centers = K, nstart = 25)`. @fig-kmeans shows the result for the geyser data.

```{r}
#| label: fig-kmeans
#| fig-cap: "K-means with K = 2 on the standardized Old Faithful data, mapped back to the original units. Crosses mark the cluster centroids."
#| fig-width: 5
#| fig-height: 3.6
set.seed(8)
km <- kmeans(scale(faithful), centers = 2, nstart = 25)
plot(faithful$eruptions, faithful$waiting, pch = c(1, 16)[km$cluster],
     xlab = "eruption duration (min)", ylab = "waiting time (min)")
cent <- aggregate(faithful, list(km$cluster), mean)
points(cent$eruptions, cent$waiting, pch = 4, cex = 2, lwd = 3)
```

A useful way to understand what K-means produces is **vector quantization**, its application to images. Treat every pixel as an observation (its colour values as features), cluster the pixels into $K$ groups, and replace each pixel by its cluster centroid: the image now contains only $K$ distinct colours and can be stored in far less space. With $K = 100$ the compressed image is hard to distinguish from the original; with $K = 5$ it is visibly posterized; with $K = 1$ it is a single grey rectangle. There is no correct $K$ here, only a trade: more clusters reconstruct the data more faithfully, and more clusters take more space to store. Reconstruction error and model complexity pull against each other, exactly as training fit and model complexity did in @sec-overfitting, and this trade returns in earnest when the number of clusters must be chosen in the model-based framework below.

::: {.callout-tip icon="false" title="Exercise 8.1"}
(a) A dataset records height in centimetres and monthly income in euros. What happens if these two features are given to `dist()` unstandardized, and why?
(b) A researcher wants to cluster daily activity curves from accelerometers, where the same activity pattern can be shifted in time between persons. Why is Euclidean distance between the raw curves a poor choice, and what kind of distance addresses this?
(c) In @fig-dendro, what would cutting the tree at height 2.5 instead of 4 have produced, and which of the two cuts is "correct"?
(d) A colleague reports "the K-means clustering" of a dataset, run once with `nstart = 1`. Name two distinct reasons the reported clusters might not be reproducible.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.1"}
(a) Income has a standard deviation in the hundreds or thousands, height in the tens, so squared income differences dominate the sum: the "clustering" is effectively a clustering on income alone. Standardization gives the features equal weight.
(b) Euclidean distance compares the curves minute by minute, so two persons doing identical things an hour apart appear maximally different. Elastic distances such as dynamic time warping align the curves before comparing them.
(c) A lower cut crosses more branches, so it produces more, smaller clusters. Neither cut is correct in any absolute sense: the tree represents the data, and the number of clusters is a choice the analyst brings to it, ideally on the grounds discussed at the end of this chapter.
(d) The random start can land the algorithm in different local minima with genuinely different partitions, and even when the partition is identical the arbitrary labels can switch, so cluster "1" need not be the same group across runs.
:::

## Two cultures, again

Before moving to the second family, it is worth noticing what kind of object K-means is, because the difference from everything in chapters 1 to 4 illustrates a cultural divide as sharp as the one @sec-explanation-prediction described. In classical statistical modelling the analyst starts from a *model*: a probabilistic description of how the data could have arisen, with parameters that are interesting *if the model holds*. Once the model is written down, how to estimate its parameters usually follows from standard statistical theory, typically by optimizing a likelihood; the computation is a consequence, not the contribution. Machine learning often works in the opposite direction. It starts from a *computation* that produces something apparently useful, K-means being a perfect example: an intuitive procedure with no probability model in sight. The questions come afterwards: what is this computation good for, and how should it be evaluated, given the task? That order of operations, compute first and justify by usefulness, was proposed as a legitimate mode of data analysis as early as Tukey's "The future of data analysis" [@tukey1962], decades before machine learning made it routine. Neither direction is superior. The model-first route buys interpretable parameters and principled inference at the price of assumptions; the algorithm-first route buys flexibility and speed at the price of not quite knowing what the output *is*. The next section takes the model-first route to the same destination, and the payoff will be that the trade-offs become visible: what K-means silently assumes can be stated exactly once a model stands behind it.

## Clustering with a model: mixtures

Suppose the height of every person on the planet were measured and plotted.^[The measurement is invented, but the rest of the section is real.] The left panel of @fig-height shows the distribution: clearly not normal, with two peaks that look like they might come from two groups. An obvious guess is that the groups are women and men. Had each person's sex been recorded, the guess could be confirmed by drawing the histogram separately per group, as in the right panel: within each sex, height is close to normally distributed, and only the combination is not. But suppose sex was not recorded. This happens constantly in research: the usability test that omitted handedness, the survey that could not ask the sensitive question, the attitude that cannot be measured directly at all. The groups are suspected but unobserved, which is to say *latent*, and recovering them from the observed data is exactly the unsupervised task.

```{r}
#| label: fig-height
#| fig-cap: "An invented population of heights. Left: the observed distribution, which is not normal. Right: the two groups that produced it, women and men, each normally distributed, with their sum as the dashed line."
#| fig-width: 8
#| fig-height: 3
x <- seq(1.40, 2.00, length.out = 400)
f_w <- dnorm(x, 1.62, 0.065); f_m <- dnorm(x, 1.77, 0.070)
mix <- 0.5 * f_w + 0.5 * f_m
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 1))
plot(x, mix, type = "l", lwd = 2, xlab = "height (m)", ylab = "density",
     main = "observed")
plot(x, 0.5 * f_w, type = "l", lwd = 2, xlab = "height (m)", ylab = "density",
     main = "the two groups", ylim = c(0, max(mix) * 1.05))
lines(x, 0.5 * f_m, lwd = 2)
lines(x, mix, lty = 2)
```

One idea that does *not* work is to split the sample on a guess about each person's sex and compute histograms within the guessed groups: without extra information the guesses will be wrong often enough to distort both histograms. What does work is a **mixture model**. Write the density of height as a weighted sum of two normal densities,

$$
p(\text{height}) = \pi_1^X\, \text{Normal}(\mu_1, \sigma_1) + (1 - \pi_1^X)\, \text{Normal}(\mu_2, \sigma_2),
$$ {#eq-mixture}

where $\pi_1^X$ is the probability that the latent group variable $X$ takes its first value, and $\mu_k, \sigma_k$ are the mean and standard deviation of height within group $k$. (The superscript $X$ is not a power; it labels the variable the probability belongs to, a notation that is standard in this literature and becomes convenient with categorical data below. The groups are numbered rather than named because the model cannot know which component is "men": component 1 could equally correspond to women, the mixture-model version of label switching.) Fitting @eq-mixture to the observed heights recovers estimates of all five parameters, and with them the right panel of @fig-height, *without sex ever having been observed*. As the sample grows, the estimates converge on the true within-group distributions.

This result is remarkable enough that its price should be stated immediately. The model *assumes* that height is normally distributed within each group, and this assumption is not verifiable from the observed data, because the observed data are exactly the mixture in which the groups are hidden. The assumption is what converts an impossible problem, unscrambling two unlabelled distributions, into a solvable one. Everything the mixture model delivers is conditional on it, a point this chapter returns to at length.

Estimation works by a procedure worth understanding, the **EM algorithm**, because its central quantity is also the model's main practical output. Start from a guess of the parameters. Given the guess, compute for each person the **posterior probability** of belonging to each group: by Bayes' rule, the probability that a person of height $h$ is in group 1 is group 1's share of the total density at $h$,

$$
\Pr(X = 1 \mid h) = \frac{\pi_1^X\, \text{Normal}(h; \mu_1, \sigma_1)}{p(h)} .
$$

Then update the parameter estimates using every person, weighted by these posteriors: a person who is probably a man contributes mostly to the male mean, but also a little to the female mean, in proportion to the uncertainty. Iterating the two steps until nothing changes maximizes the likelihood. The posterior probabilities are the model's answer to "who is in which group": not a hard assignment but a probability per person, acknowledging that group membership is genuinely uncertain. A small computation makes this concrete:

```{r}
#| label: posterior
p_m <- 0.5 * dnorm(1.72, 1.77, 0.070)   # component density: men, at 1.72 m
p_w <- 0.5 * dnorm(1.72, 1.62, 0.065)   # component density: women, at 1.72 m
round(p_m / (p_m + p_w), 2)             # posterior probability of being a man
```

A person of 1.72 metres is more likely a man than a woman, but far from certainly, and the posterior of 0.70 records exactly how far. Where the two component curves cross, the posterior is 0.50 and the observation contributes equally to both groups.

### Latent profile analysis: mixtures of several continuous variables

With one observed variable the price of the assumption was steep. With several variables the model becomes both more useful and more credible, because the variables corroborate each other. The multivariate version of @eq-mixture, a mixture of multivariate normal distributions, is called a **Gaussian mixture model** in machine learning and **latent profile analysis (LPA)** in the social sciences; the two names refer to the same model. Each component $k$ now has a mean *vector* and a covariance *matrix*, so in two dimensions a component is an ellipse: its centre is the mean vector, its size, shape, and orientation are the covariance matrix.

The covariance matrices can be constrained, and the constraints matter. Requiring all components to share one covariance matrix gives clusters of equal size and shape; letting every component have its own gives full flexibility at the price of many more parameters, since a covariance matrix in $q$ dimensions has $q(q+1)/2$ of them. This is the bias-variance trade-off of @eq-bias-variance in new clothing: a rigid parameterization can miss the clusters' true shapes, a flexible one can chase noise. The `mclust` package [@scrucca2016] organizes these options in a three-letter code for volume, shape, and orientation, each either **E**qual across components, **V**ariable, or **I** (spherical): "EII" is equal-volume spheres, "EEE" is shared ellipses, "VVV" is every-component-for-itself. By default `Mclust` fits every parameterization for a range of component counts and reports the best model, chosen by a criterion the next section examines closely:

```{r}
#| label: mclust-faithful
#| message: false
library(mclust)
mc <- Mclust(faithful, verbose = FALSE)
summary(mc)
```

```{r}
#| label: fig-mclust
#| fig-cap: "Left: BIC for the mclust parameterizations on the Old Faithful data, by number of components; higher is better in mclust's sign convention. Right: the selected model's classification, with the component ellipses drawn."
#| fig-width: 8
#| fig-height: 3.4
#| echo: false
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 1))
plot(mc, what = "BIC")
plot(mc, what = "classification", xlab = "eruption duration (min)",
     ylab = "waiting time (min)")
```

@fig-mclust shows the criterion behind that choice and the fitted components. Two details of the output deserve attention. First, K-means turns out to be hiding inside this model: constraining the components to equal-sized spheres (EII), the class proportions to be equal, and the posterior probabilities to hard 0/1 assignments makes model-based clustering computationally equivalent to K-means. The algorithm of the previous section was a mixture model all along, with assumptions nobody had to state because no model was written down; spelled out, they are strong ones, spherical clusters of equal size, and @fig-faithful's elongated clusters violate them visibly. Second, the selected solution has *three* components where the eye insists on two clusters. This is not a mistake. The upper-right cluster is not elliptical, so the model uses two overlapping ellipses to describe its shape, and the mixture describes the *density* correctly at the cost of the component–cluster correspondence. When the groups themselves are the quantity of interest, two repairs exist: select the number of components by a criterion that rewards well-separated components (the ICL below), or fit the well-fitting mixture first and then **merge** components that overlap heavily into single clusters [@baudry2010], as `mclust::clustCombi()` does, which reduces these three components to the two clusters the picture shows. The distinction between *components*, the normal densities the model uses, and *clusters*, the groups the researcher means, is worth keeping.

### Latent class analysis: mixtures of categorical variables

When the observed variables are categorical, the within-class normal distributions are replaced by within-class category probabilities, and the model is called **latent class analysis (LCA)**. It is the discrete-observed, discrete-latent cell of @tbl-lvm, which organizes several earlier chapters' methods at once:

|                        | Latent variable continuous | Latent variable discrete |
|------------------------|---------------------------|--------------------------|
| **Observed continuous** | factor analysis           | latent profile analysis  |
| **Observed discrete**   | item response theory      | latent class analysis    |

: Names of latent variable models, by the type of the observed and latent variables. {#tbl-lvm}

For three categorical observed variables $A$, $B$, $C$, the latent class model says

$$
\pi_{abc}^{ABC} \;=\; \sum_{x} \pi_x^X \, \pi_a^{A \mid X} \, \pi_b^{B \mid X} \, \pi_c^{C \mid X} ,
$$ {#eq-lca}

read as: the probability of observing the response pattern $(a, b, c)$ is a weighted sum over the latent classes $x$ of the probability of that pattern within each class, weighted by the class sizes $\pi_x^X$. The crucial structure sits in the product $\pi_a^{A \mid X} \pi_b^{B \mid X} \pi_c^{C \mid X}$: *within* a class, the observed variables are independent. This assumption, **local independence**, is LCA's version of the within-class normality assumption above, and it carries the same double role: it is what identifies the model, and it is what the observed data cannot fully check. It also supplies the model's interpretation. The classes are constructed exactly such that they explain the associations among the observed variables: people are similar within classes and different between them, and whatever relationships the variables show in the whole sample should vanish once class membership is held constant.

The LISS panel provides a natural application. The health module asks every adult respondent whether a physician has ever diagnosed each of a list of conditions. Ten of these diagnoses, from angina to cancer, are binary indicators of a respondent's pattern of chronic disease, and a standing question in health research is whether such conditions cluster into recognizable **multimorbidity** patterns: combinations that co-occur in the same people, such as the cardiometabolic cluster of hypertension, high cholesterol, and diabetes. The script `analysis/liss/06-prepare-lca.R` builds the dataset: 4,647 respondents with complete diagnosis batteries. The model is fitted with the `poLCA` package [@linzer2011].

```{r, cache=TRUE}
#| label: lca-fits
#| message: false
#| results: hide
library(poLCA)
d <- readRDS("../data/liss-health-social-integration/liss_lca_conditions.rds")
items <- c("angina", "heartattack", "hypertension", "cholesterol", "stroke",
           "diabetes", "lung", "asthma", "arthritis", "cancer")
dd <- d[items] + 1                       # poLCA codes categories 1, 2
f  <- as.formula(paste("cbind(", paste(items, collapse = ","), ") ~ 1"))
set.seed(8)
fits <- lapply(1:5, function(k)
  poLCA(f, dd, nclass = k, nrep = if (k == 1) 1 else 10,
        maxiter = 3000, verbose = FALSE))
```

```{r}
#| label: lca-table
round(t(sapply(fits, function(m)
  c(classes = length(m$P), loglik = m$llik, npar = m$npar,
    AIC = m$aic, BIC = m$bic))), 1)
```

The `nrep = 10` argument runs the EM algorithm from ten random starting points per model and keeps the best, for the local-maxima reason K-means already taught; skipping it is the most common error in applied LCA. Each added class costs eleven parameters, one class proportion and ten conditional probabilities, and buys some log-likelihood. The BIC is smallest for the three-class model, while the AIC is still falling at five classes, a disagreement to which the next section returns. Staying with the BIC's choice, @fig-lca-profile displays the three-class solution as a **profile plot**: one line per class, showing the probability of each diagnosis within that class.

```{r}
#| label: fig-lca-profile
#| fig-cap: "Profile plot of the three-class latent class model of ten physician diagnoses, LISS 2025 (n = 4,647). Lines show the probability of each diagnosis within each class; percentages are the estimated class sizes."
#| fig-width: 7.5
#| fig-height: 3.6
#| echo: false
m3 <- fits[[3]]
pr <- sapply(m3$probs, function(p) p[, 2])
ord <- order(m3$P, decreasing = TRUE)         # stabilize label order by size
pr <- pr[ord, ]; sizes <- m3$P[ord]
par(mar = c(6.5, 4, 1, 8))
matplot(t(pr), type = "b", pch = c(16, 17, 15), lty = 1, lwd = 1.5,
        xaxt = "n", xlab = "", ylab = "P(diagnosis | class)", ylim = c(0, 0.7))
axis(1, at = seq_along(items), labels = items, las = 2, cex.axis = 0.8)
legend(x = length(items) + 0.6, y = 0.7, xpd = NA, bty = "n", cex = 0.8,
       pch = c(16, 17, 15), col = 1:3, lty = 1,
       legend = sprintf("class %d (%.0f%%)", 1:3, 100 * sizes))
```

The classes are recognizable. The largest, at about three quarters of the sample, is a *healthy* class in which every diagnosis is rare. The second, close to a fifth of respondents, is a *cardiometabolic* class: hypertension in about six of ten members, high cholesterol in more than half, elevated diabetes and arthritis. The smallest, around five percent, is a *respiratory–musculoskeletal* class in which asthma, chronic lung disease, and arthritis are the distinguishing diagnoses. Each respondent additionally receives a posterior probability of belonging to each class, exactly as in the height example, and these posteriors carry real uncertainty: on average the largest posterior is well below 1. That uncertainty must be respected when the classes are used further. Assigning each person to their most likely class and then regressing that assignment on covariates treats the classification as data when it is an estimate, and biases the resulting associations toward zero; corrected **three-step** procedures exist for exactly this task [@bakk2013].

The local independence assumption is not entirely beyond checking. Within each class, pairs of indicators should be unrelated; computing the residual association of each pair within classes, for instance as a chi-square in the class-specific cross-table (a **bivariate residual**), points to pairs the model fails to explain. A large residual between asthma and chronic lung disease, say, would signal that these two diagnoses share something beyond the classes, and a model treating them jointly, or an extra class, may be needed. What such checks can never do is validate local independence wholesale, for the reason the height example made plain: the classes are latent, so the full pattern of within-class association is not observable.

::: {.callout-tip icon="false" title="Exercise 8.2"}
(a) In @eq-lca, how many parameters does a $K$-class model for ten binary indicators have, and how many does the saturated model for the $2^{10}$ response patterns have? For which $K$ does the first exceed the second?
(b) State the three restrictions that turn latent profile analysis into K-means.
(c) A colleague fits the three-class model above, assigns every respondent to their most likely class, and reports "class membership predicts mortality". Name the two distinct sources of overconfidence in that sentence.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.2"}
(a) The $K$-class model has $(K - 1) + 10K$ parameters: $K-1$ free class proportions and ten conditional probabilities per class. The saturated model has $2^{10} - 1 = 1023$. The model reaches 1023 parameters at $K = 93$, but it stops being identified far earlier, as the next section discusses: counting parameters is only the crudest identification check.
(b) Equal-sized spherical covariance matrices in every class (EII), equal class proportions, and hard assignment of each observation to one class (posterior probabilities forced to 0 or 1).
(c) The classification is an estimate with uncertainty, and treating the modal assignment as observed data biases the association toward zero, which the three-step corrections repair. And the classes themselves are model-dependent: as the next section argues, a well-fitting mixture does not establish that these particular groups exist, so "predicts mortality" inherits every assumption behind the class solution.
:::

## Choosing K, and trusting the result

Everything above postponed the same decision: how many clusters or classes, and which parameterization. In supervised learning this kind of decision went to the validation set, and the test set stood ready to price the final choice honestly. Here there is no outcome to score against, which is why model selection and evaluation deserve more space in this chapter than in any other, and why this section is the chapter's real destination. The available tools form a menu, and knowing what each item does, and cannot do, matters more than any single recipe.

**Information criteria.** For model-based clustering the standard tools are the same AIC and BIC that appeared in earlier chapters,

$$
\text{AIC} = -2\,\ell + 2\,p, \qquad
\text{BIC} = -2\,\ell + p \log n ,
$$

where $\ell$ is the maximized log-likelihood, $p$ the number of parameters, and $n$ the sample size. Their logic is precisely the logic of chapters 6 and 7. The in-sample log-likelihood, like the training error, flatters every added class, so a criterion must correct it, and what AIC estimates is exactly *what the out-of-sample log-likelihood would have been*: the $2p$ term is an estimate of the optimism of the in-sample fit, so that comparing AICs is an approximation to comparing models on fresh data, the very quantity chapters 6 and 7 measured by holding data out [@hastie2009]. The BIC's stronger penalty, $p \log n$, approximates a Bayesian comparison of the models and, in mixture models, tends to select fewer, better-separated classes; it is the default in this literature and it is consistent, selecting the true number of components when one exists and $n$ grows. The vector-quantization trade of this chapter's first half is the same computation in different words: the log-likelihood is the reconstruction, the penalty is the file size. And because the criteria are estimates of out-of-sample performance, the direct route is also available: the log-likelihood can simply be cross-validated, scoring each held-out fold under the model fitted to the rest, with all of @sec-evaluation's machinery and cautions applying unchanged.

The two criteria disagreed above, BIC choosing three classes and AIC five, and such disagreement is typical rather than exceptional: AIC aims at predictive accuracy and will add classes that improve the density estimate slightly, BIC aims at recovering a parsimonious true structure. There is no neutral arbiter, because the criteria answer different questions, and the choice between them is a choice about the task. When the classes are to be interpreted and used, parsimony and separation usually matter more than the last increment of fit, which favours the BIC, possibly alongside the **ICL**, a BIC variant that additionally penalizes solutions whose classes overlap so much that observations cannot be confidently assigned [@biernacki2000]. Likelihood-ratio testing of $K$ against $K+1$ classes is also possible, but the standard chi-square theory fails at this boundary, and the accepted replacement is the bootstrapped likelihood-ratio test, which simulates the test's distribution under the smaller model [@nylund2007].

**Internal indices, stability, and external checks.** For algorithmic clustering without a likelihood, and as complements beside it, three further families exist. Internal validation indices score a clustering from the data alone, the best known being the **average silhouette width**, which compares each observation's distance to its own cluster with its distance to the nearest other cluster [@rousseeuw1987], and the **gap statistic**, which compares the clustering criterion against its expectation under no clustering [@tibshirani2001gap]. **Stability assessment** asks a question closer to this book's heart: would the clusters survive a replication? Reclustering bootstrap resamples and measuring how often each cluster recurs, for instance by the Jaccard index as in `fpc::clusterboot()`, exposes clusters that are artifacts of the particular sample [@hennig2007]; the same logic applies to perturbing the hyperparameters, the linkage, the distance, the parameterization, and checking whether the substantive story survives. And **external validation** checks the clusters against information not used to form them: if the multimorbidity classes above are real, they should differ in age, subsequent health care use, and mortality, in sensible directions. None of these indices knows the researcher's aim, however, and the choice among clusterings ultimately depends on what the clustering is *for* [@hennig2015]: a market segmentation is good when the segments respond differently to actions, a multimorbidity typology when it organizes prognosis and treatment, and no index measures either.

**Identification, and its empirical failure.** With latent variables a question precedes model selection: is the model identified at all? @sec-identification introduced identification for causal quantities; the parametric version asks whether different parameter values could produce the same distribution of the observed data. For mixtures the answer degrades gracefully and dangerously. A latent class model can be formally underidentified, with strictly more parameters than the observed table can support, the parameter count of Exercise 8.2 being only the crudest bound [@goodman1974]. More insidious is **empirical underidentification**: the model is identified on paper, but the observed data barely constrain some parameters, so the likelihood surface is nearly flat in some directions. LCA reaches this state quickly, especially with rare indicators, small classes, and generous $K$. Its symptoms are visible to an analyst who looks. Repeated estimation from many random starts returns different solutions with practically equal log-likelihoods; estimated probabilities land on the boundary at exactly 0 or 1; some class sizes collapse toward zero; and standard errors of some parameters explode. The six-class model for the LISS diagnoses shows the syndrome on demand:

```{r}
#| label: underidentification
#| message: false
sixes <- t(sapply(1:8, function(s) {
  set.seed(100 + s)
  m <- poLCA(f, dd, nclass = 6, nrep = 1, maxiter = 3000, verbose = FALSE)
  c(loglik = m$llik, smallest_class = min(m$P),
    boundary_estimates = sum(unlist(m$probs) < 1e-4 |
                             unlist(m$probs) > 1 - 1e-4))
}))
round(sixes, 2)
```

Eight starts produce several distinct maxima whose log-likelihoods differ only in the first decimal, each with a dozen or more boundary estimates and smallest classes holding at most a few percent of the sample. No one of these solutions deserves the name "the six-class model": the data cannot tell them apart, and any substantive story told about one of them would have come out differently on the next start. The practical defences follow from the symptoms: always use many random starts and inspect the spread of the maxima, distrust boundary estimates and near-empty classes, and when the syndrome appears, reduce $K$, restrict parameters, or collect more indicators, since each added indicator multiplies the observable information while adding only $K$ parameters.

**Equivalence, one more time.** Even a well-identified, well-fitting mixture supports a weaker conclusion than it appears to. Very different models can fit the same data equally well: a heavily restricted model with many classes can reproduce the observed distribution as well as a flexible model with few classes, just as the three-component mclust solution and the two merged clusters described the Old Faithful density equally well. This is the equivalence problem of @sec-theory-to-glm and @sec-identification returning in mixture clothing, and it has the same consequence. Fit cannot distinguish between the equivalent descriptions, so the choice among them is made by assumptions, and therefore the inference "the model fits, hence these are the true latent groups" is never available. What is available is the conditional statement this book has practiced since @sec-identification: *if* the population consists of groups within which the indicators are independent (or normal), *then* the data point to these groups, with this many classes, of these sizes. The assumptions carry the conclusion, and for latent variables some of them are uncheckable in principle, since checking would require observing the very variable that is latent.

Uncheckable is not the same as unsupportable, and here the book's oldest theme returns: design. Local independence is an assumption about *why the indicators are correlated*: it says the latent class is the only common cause, so that within a class the remaining variation in each indicator is noise specific to that indicator. Whether that is plausible depends on where the indicators come from. Ten self-reported diagnoses from a single questionnaire share a respondent, a sitting, a memory, and a reporting style, and any of these can correlate the indicators within classes, violating local independence in ways no fit statistic will reliably flag. Now consider instead a design in which one indicator of, say, problematic debt comes from a survey question, a second from an administrative register of payment arrears, and a third from a digital trace of account activity. These three measurements share essentially one thing: the person's true state. Their error sources, self-presentation, administrative lag, algorithmic misclassification, are distinct by construction, so conditional independence given the latent state is not a hope but a designed property. The same assumption that is heroic within a single survey becomes credible across sources, and nothing about the model changed, only the design behind its inputs. Latent variable models do not remove the need for input thinking; they relocate it, from the interpretation of coefficients to the provenance of indicators.

::: {.callout-tip icon="false" title="Exercise 8.3"}
Using the AIC/BIC table for the LISS diagnoses:

(a) Reconstruct the BIC of the two-class model from its log-likelihood and parameter count, with $n = 4{,}647$.
(b) The AIC still prefers five classes. Explain, in the vocabulary of chapters 6 and 7, what each criterion is trying to estimate, and construct a scenario in which the five-class model is the right choice after all.
(c) Propose a cross-validation scheme that would compare the three- and five-class models directly on out-of-sample performance, and state what quantity is scored on the held-out folds.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.3"}
(a) $\text{BIC} = -2 \times (-9361.1) + 21 \times \log 4647 = 18722.2 + 21 \times 8.44 \approx 18899.5$, matching the table.
(b) The AIC estimates out-of-sample log-likelihood, so it rewards any structure that helps describe fresh data, including small or overlapping components; the BIC approximates the evidence for the model and demands more per parameter, favouring few well-separated classes. If the task were density estimation, generating realistic synthetic response patterns, say, the five-class model's better predictive fit is exactly what is wanted. For interpretable groups, the three-class solution's separation matters more.
(c) Split respondents into folds; fit each candidate on the training folds, with many random starts; compute each held-out respondent's log-likelihood under the fitted model, summing over classes as in @eq-lca; average per fold and compare. The scored quantity is the out-of-sample log-likelihood that AIC approximates, and everything @sec-evaluation said about folds and leakage applies, including refitting the entire pipeline within each fold.
:::

## What the missing judge means

This chapter closes the machine learning part of the book, and it ends on a reversal worth stating plainly. Chapters 6 and 7 built their case on output validity: fix the task and the metric, and models can be compared empirically, with assumptions demoted to whatever helps prediction. That case rested entirely on the existence of a supervisor, an observed outcome against which performance could be measured honestly. Unsupervised learning has no such outcome, and the consequence runs through everything in this chapter. The number of clusters is not measurable but chosen, with criteria that answer different questions; the clusters themselves are conditional on assumptions, some uncheckable in principle; equivalent models tell different stories about the same fit; and the strongest support available comes not from any statistic but from design, in the provenance of the indicators and in validation against external information. In other words: where the supervisor leaves, input validity returns. The questions of chapters 1 to 4, what was measured, how, on whom, and what must be assumed, were never machine learning's rivals; they are what machine learning stands on whenever the ground of honest output measurement is unavailable. A researcher who can run both playbooks, and knows which one the problem calls for, has the whole book.

Further reading: chapter 12 of @james2021 covers clustering from the machine learning side; @oberski2016 is a compact introduction to latent profile and latent class analysis from the social science side, on which this chapter's treatment of mixtures draws; @hennig2015 is the standard guide to choosing a clustering strategy; and @mclachlan2000 is the comprehensive reference on finite mixture models.

## Test yourself {.unnumbered}

::: {.callout-tip icon="false" title="Exercise 8.A"}
A municipal health service wants to identify types of problematic alcohol use among 12,000 residents, combining three indicators: a survey self-report, emergency-department records, and pharmacy dispensing data.

(a) Say why this three-source design supports the local independence assumption better than three survey questions would, and name one residual threat to it.
(b) Specify a defensible procedure for choosing the number of classes, naming at least three distinct tools and the order in which you would consult them.
(c) The four-class solution includes a class of 1.5 percent of residents with two conditional probabilities estimated at exactly zero. What do you suspect, and what three checks would you run?
(d) The service wants to target an intervention at the highest-risk class. Which chapters' machinery does that decision invoke beyond this one's, and why?
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.A"}
(a) The three sources share the resident's true state but little else: survey errors come from self-presentation and recall, ED records from presentation and coding practices, dispensing data from prescription behaviour. Distinct error mechanisms make within-class independence a designed property rather than an assumption of convenience. A residual threat: all three sources may miss the same people, for example residents who avoid care entirely, which correlates the indicators within the healthy class.
(b) Fit $K = 1, 2, \ldots$ with many random starts; consult the BIC (and ICL) for a defensible range; run the bootstrapped likelihood-ratio test on the boundary candidates; check stability by refitting on bootstrap resamples; and weigh interpretability and class separation for the final choice, documenting the whole path rather than the winner alone.
(c) Empirical underidentification. Check the spread of log-likelihoods across many starts; inspect standard errors around the boundary estimates; and refit with $K - 1$ classes and with the boundary parameters restricted, comparing fit.
(d) Targeting is an intervention, so its justification is causal: the designs of @sec-rq-to-interpretation and the identification reasoning of @sec-identification. That the class predicts harm is an umbrella claim; that intervening on its members reduces harm is a rain-dance claim, in the vocabulary of @sec-explanation-prediction.
:::

::: {.callout-tip icon="false" title="Exercise 8.B"}
Two published analyses of the same national survey report latent class models of adolescent risk behaviour. Study 1 reports a four-class model with unrestricted conditional probabilities. Study 2 reports a seven-class model in which many conditional probabilities are constrained equal across classes. Both report excellent and nearly identical fit, and each claims to have found "the" typology.

(a) Explain why the near-identical fit is not surprising, using this chapter's vocabulary and its antecedents in earlier chapters.
(b) Neither study used multiple random starts. What specifically is at risk in each?
(c) A reader asks which typology is true. Write the two-sentence honest answer.
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.B"}
(a) The two models are close to equivalent: restrictions traded against classes can leave the implied distribution of the observed responses nearly unchanged, so fit cannot separate them, just as equivalent path models fit identically in @sec-identification. The data constrain the observed distribution; the partition of that distribution into classes comes from the assumptions.
(b) Each reported solution may be a local maximum, so the published parameters, and with them the substantive typology, might not even be the best solution of the authors' own model; and with seven classes, Study 2 is additionally at high risk of empirical underidentification, with several near-equal maxima describing different typologies.
(c) Neither is known to be true: each is the best description of the data conditional on its own restrictions, and the data cannot adjudicate between the restrictions. What can be reported honestly is the conditional claim, together with checks, stability, external validation, design of the indicators, that make one set of assumptions more defensible than the other.
:::

::: {.callout-tip icon="false" title="Exercise 8.C"}
Return to the Old Faithful data.

(a) K-means with $K = 2$ and the merged mclust solution produce nearly the same two clusters. State one substantive conclusion the mixture model licenses that K-means does not, and why.
(b) mclust's chosen model used three components for two clusters. Explain to a non-statistician colleague why "the algorithm found three groups" would be a misreading.
(c) Suppose eruptions were recorded by two different instruments, and the instrument identity were unavailable. What syndrome from this chapter should the analyst worry about, and which checks apply?
:::

::: {.callout-note collapse="true" icon="false" title="Answer 8.C"}
(a) The mixture model yields posterior probabilities, so it can report that eruptions near the boundary belong to either type with stated uncertainty, and it estimates the types' proportions with standard errors; K-means's hard assignments assert certainty the data do not contain.
(b) The three normal components are the model's building blocks for describing the shape of the data's density, not necessarily the groups of interest: one banana-shaped cluster needs two ellipses to trace it. Components are vocabulary; clusters are the claim.
(c) An unmeasured grouping variable, the instrument, can generate spurious classes or distort real ones, and within-"type" correlation induced by the instrument violates local independence. Stability checks across subsamples, external validation against any available metadata, and scrutiny of whether extra small classes appear with boundary estimates are the applicable defences.
:::
